### 2022-05-26 Manuscript/Manuscript.tex
Update after feedback

\end{flushleft}
\section*{
Abstract
}
The skill to recognize and classify taxa requires knowledge that is less and less common in the scientific community. At the same time, it is clear that these skills are needed in biodiversity monitoring for management and conservation, especially when carried out by citizen scientists. Formalizing the required knowledge in digital identification keys is a way of making such knowledge more available for professional and amateur observers of biodiversity. In this paper we describe Clavis, a modern open framework for capturing knowledge required for taxon identification through digital keys, allowing for a level of detail beyond that of any current format. We exemplify each concept using Pokémon as a fictional taxonomic group.
\section*{
Introduction
}
Distinguishing biological taxa from one another is a necessity in biodiversity monitoring. As species are going extinct at an unprecedented rate~\cite{Ceballos2015, Johnson2017}, we need to monitor as much as we can to identify statuses and trends. Meanwhile, research and management is currently facing a ``taxonomic impediment'', where taxonomic knowledge is disappearing from the scientific community~\cite{Engel2021}.

Paradoxically, while taxonomic expertise is becoming scarce, species observations are being reported like never before. The bulk of the observational data currently available originates from citizen science (in which observations are made by non-professional volunteers~\cite{Silvertown2009}). There are, however, large taxonomic biases in the data collected, especially within citizen science. For example, of publicly available data from the Global Biodiversity Information Facility (GBIF, the largest aggregator of biodiversity observations in the world~\cite{GBIFhomepage}), some 67\% of over 2 billion observations pertain to birds~\cite{GBIF_data_taxonomy}. This means that less charismatic species, and species that are more difficult to identify, generally remain severely under-reported~\cite{Troudet2017}. Many such species play important ecological roles or can serve as indicators of broader ecosystem health, but the fact that few observers are able to reliably identify these species leads to their underreporting, obscuring important trends from the view of science and management~\cite{Dobson2020}.

The knowledge needed to identify species resides in large part in the heads of taxonomic experts. This invaluable experience is slowly disappearing however, as newly educated biologists are, to an increasing extent, trained with focus on skills in genetics, data management, and biodiversity informatics rather than traditional taxonomy. This training, and an academic reality of shorter periods with funding and more temporary contracts, gives fewer researchers the opportunity to invest the years of experience needed to become truly knowledgeable with regards to any substantial taxon, as was more commonplace in the past~\cite{Ebach2011}.

Identification keys are one of the most important tools used by taxonomists to identify taxa, and thus also to share with others the information needed for distinguishing taxa. No taxonomist studying a large taxon in depth can get around identification keys, and keys are often the starting point when trying to learn a new organism group. Identification keys can, however, be quite challenging to use, especially for novel users, and their use often represents a significant barrier for the prospective learner. For this and other reasons, experience with identification keys is something receiving less focus in most current biology curricula. Furthermore, as taxonomic knowledge is becoming increasingly fragmented, existing identification keys in literature are gradually becoming outdated as taxonomies are changing according to new insights.

These issues are exacerbated in the context of citizen science. Whereas prospective taxonomists have a professional incentive to learn to use an identification key, a steep learning curve can quickly put off the more casual citizen scientist. With its unrivaled amounts of data, but also added concerns regarding the reliability of the taxon identifications~\cite{Crall2011,Burgess2017} and the pronounced over-representation of more charismatic taxa~\cite{Callaghan2020,Bayraktarov2019,Boakes2016}, citizen science stands especially much to gain from good identification keys with a low usability threshold.

Digital identification keys can address these challenges in a number of ways. The formalized and machine-readable way in which the data are stored allows for easier revision, as well as multiple options for their display to the end user, tailored to their level of experience. Digital keys allow for the representation of more complex relationships between species' characteristics and identifications inherent to biological complexity. These relationships cannot as easily be represented in a linear form as is used in traditional, paper-based keys. With a suitable interface, they open up the cumulative experience from taxonomic experts to a broader public like citizen scientists, while potentially being more reliable and educational than other identification methods such as automated image recognition. In contrast to paper-based keys, digital keys can also be combined with one another, integrating the information from several separate keys into one seamless user experience. And conversely, a digital key can easily be limited to just a subset of the taxa in it, for instance only taxa belonging to a certain habitat or recorded in a certain geographic range. Another benefit is the option of saving the user input and the key together with the observational data, as a means of transparent identification that can be reviewed at any time in the future for quality control or in the case of new taxonomical insights.

Having a fully open and well-defined, platform-independent format is essential in ensuring that identification keys and the tools needed to display and create these are and remain both interoperable, interpretable across platforms, and freely available to all. A number of digital identification key formats exist~\cite{Dallwitz1980, Lucid}, but these come with a number of caveats in terms of limitations in what they can represent, ease of use, openness, etc. A well-defined format addressing this alleviates the technical burden in capturing taxonomic knowledge for future use.

In this manuscript we document Clavis; an open format for identification keys, aiming to cover the aforementioned requirements in an open and lightweight manner, and serving as an example of a way to store and exchange crucial taxonomic knowledge. Clavis is intended to capture all the requirements of traditional keys while adding the flexibility and complexity possible with digital keys. This means that the format is well suited for representing most, if not all, existing keys, both paper based and digital, as well as for designing new keys. The name Clavis means ``key'' in Latin, and is a recursive acronym for \underline{C}lavis \underline{L}ightweight \underline{A}nd \underline{V}ersatile \underline{I}dentification \underline{S}chema. This article describes the Clavis format itself. Example code handling the business logic of keys described using Clavis will be published separately at a later time. Descriptions of each of Clavis' data types and emerging properties are illustrated here using fictional taxa that are discretely defined and not subject to taxonomic debate or change, while exhibiting the required complexity for a demonstration of the more intricate features of the format. Our fictional creatures of choice for these examples are a selection of Pokémon~\cite{Pokemon_wiki}.

\section*{
Material and methods
}
Identification keys exist in many forms, but all are built on the same basic principle. The user is asked a number of questions about the entity being identified, and by answering these questions, more and more taxa are excluded from consideration until (ideally) only one taxon remains, which is the result of the identification process.

Traditional single access keys consist of a decision tree with a single fixed path from beginning to end for each possible outcome. On choosing an answer from the alternatives for each of the questions, one is directed to the next question to be answered, until one ends up with a result taxon. Such keys can be either dichotomous, meaning that each question always has only two alternatives, leading to a bifurcating decision tree, or they can be multichotomous, meaning that each question may have more than two alternatives. One upside of single access keys is that they are easy to represent on paper. A drawback is that a user has to go through all questions in a specific order, having one and only one question to answer at any given time. This means that one cannot use only part of the key for a subset of taxa, nor easily go back and redo a question. Also, if a question cannot be answered, there is no way to proceed.

An alternative approach is the multiple access key. Such keys are typically stored as matrices where taxa and questions (characters) are stored as rows and columns in a tabular format, with cell values linking the taxa to their characters. Each character may have two or more possible answers (states), or be defined as a numerical value or range of values. This allows for a selection of taxa to be made that one wishes to distinguish between, rather than having to traverse the entire key. In a fully populated matrix key, every character is assigned a value (i.e. is scored) for every taxon. The user then has access to all the questions/characters at once and is not required to answer them in any particular order. Choosing any alternative for any of the questions will exclude from consideration all those taxa that are not compatible with that alternative, bringing the user a step closer to the answer. The upside of this approach is that the user has several paths to the answer, and can choose to avoid questions that are difficult or impossible to answer. The downside is that the number of choices can easily be overwhelming, and it is generally left to the user to try to choose the best path. Also, in real life situations it is very rare to find a set of characters that are both possible to score in a meaningful way for all the taxa in the key, and sufficient for distinguishing between them. More often, some characters will be possible to score for all taxa, whereas others may be essential for distinguishing certain taxa in the key, yet be totally inapplicable to others.

To address this, a matrix can also be made to be sparse, meaning that not all characters are scored for all taxa. The interface can then display only those characters that have been scored for all the taxa under consideration. As taxa are being excluded by the user’s choices, further characters that are scored for all the remaining taxa will become available. With this approach, it is possible to make a matrix key that behaves equivalently to a single access key. But the key can also be made to contain several distinguishing characters for any set of taxa, so that the user has more than one choice at each step and is thus not restricted by having to answer only one question at a time in a specific order.

The sparse matrix approach is more powerful than both the single access and the full matrix approach, in that both of the latter approaches are subsets of what the sparse matrix can represent. Still, there is much information one would ideally like to store about taxa, characters, and the relationships between the two, which cannot be easily represented in a tabular format. For instance, the tabular format is not well suited for dealing with polymorphism in a straightforward way, i.e. the situation where a taxon can have more than one possible value for a given character. Also, it may often be desirable to treat taxa hierarchically rather than as a simple list of equally ranked entities, something which is difficult to do in the matrix format. Thus, representing taxonomic knowledge in this tabular format restricts the complexity of the information that can be stored.

The method of storing and exchanging taxonomic knowledge described here, is developed as a non-tabular multiple access key. The core of any Clavis key is a collection of statements. A statement links a taxon to a character, and specifies the state or numerical value the taxon has for that character (see Fig.~\ref{fig1}). E.g. the color (character) of species x (taxon) is red (state), or the length in mm (character) of genus x (taxon) is 4-11 (numerical range).

\begin{figure}[!h]
  \includegraphics[width=\textwidth]{Images/Fig1}
  \caption{{\bf The core concepts of any Clavis key.}
  \underline{Taxa} (left) and \underline{characters} (right) are connected through \underline{statements} (center). Each statement refers to one taxon and one character, linking the two by a value, usually one of the possible \underline{states} of the character, or a numerical value.}
  \label{fig1}
\end{figure}

Additional information can be linked to taxa, characters, states and statements in a number of ways, capturing knowledge and relationships that cannot be represented in a tabular form in a practical way. Such information includes media elements, textual descriptions, and contextual information such as geographic scope. The Clavis format incorporates all the possibilities of the previously described approaches, so that any single access or matrix key can be transcribed to it.

Clavis compliant keys are written in JSON (JavaScript Object Notation)~\cite{JSON}, which has become the \textit{de facto} standard for data exchange over the internet, as it is an open, flexible format with widespread use and support within all modern programming languages. Clavis itself is a JSON-schema~\cite{JSON_schema}, a formal definition of the structure and content of a valid Clavis JSON file. JSON-schemas are machine readable and can be used to automatically validate the compliance of a JSON file, highlighting any issues that need to be solved. While possible, it is generally not practical to manually write the JSON of a Clavis-compliant key. In order to facilitate key creation and editing, one would generally provide taxonomists with a key editing interface that lets them easily record their knowledge, and let the key editor generate the JSON. The latest version of Clavis can be found at \url{https://github.com/Artsdatabanken/Clavis}.

\subsection*{
Implementation
}
As an exchange and storage format, Clavis does not dictate how it is to be implemented. Different interfaces can implement it differently, depending on the purpose of the interface and the user group. To this end, interfaces serving to edit or display may disregard certain non-essential functionality supported by Clavis. One could for example create a key editor not supporting media files or multilingualism. Other aspects of the format are crucial however, and require support and unambiguous interpretation.

Identification is a matter of excluding taxa, and is done by letting the user select a state or numerical value for characters. Each time the user provides a new fact in this way, all taxa with conflicting statements are excluded (see Fig.~\ref{fig2}). The user can also be given the option to exclude a state rather than affirm one if there are more than two states to choose from. Characters that have not been answered but only have a single possible answer remaining for all remaining taxa can be hidden.

@@

Whenever taxa are excluded, characters that were previously hidden may now have become relevant if all the remaining taxa have a statement pertaining to it, as illustrated in Fig.~\ref{fig2}. It is important to check for states that, while not having been answered directly by the user, can be deduced from the statements related to the remaining taxa: if all the remaining taxa are known not to have a certain state, that state can be disabled. In such cases, the state still needs to be visible to the user as it provides context to the remaining options, but it should not be possible for the user to select it. In all cases, only characters that are linked by statements to \underline{all} the currently non-excluded taxa should be shown to the user.

Things become more complicated when allowing the user to undo previous answers. Undoing an answer might render a character that was subsequently answered irrelevant again. To avoid answers to this now irrelevant character from eliminating taxa as possibilities, an implementation needs to be able to remove or ignore such answers on characters that were rendered irrelevant.

@@

To illustrate the features supported by the Clavis format, we here describe features and give some example uses. The examples relate to fictional creatures that are not subject to taxonomic debate or change. For this we chose Pokémon as they appear in the mobile game Pokémon Go, as these are clearly defined, yet exhibit the complexity required to demonstrate the various features of the format. Keys for natural taxa would rarely use as much of the possible functionality as we demonstrate here in a single key. The key is valid in accordance with the JSON-schema, but it does not contain all the information needed to distinguish between all of the Pokémon mentioned in it. We also provide an example of a non-artificial key that does identify all taxa in it, but without aiming to demonstrate all possible features that Clavis supports. This key covers all species of titmice in Norway and can be found in the supplementary material.

@@

Polygon of where the key is valid

@@

The language(s) the key supports. Texts for multiple languages in other key objects are in the order defined here

@@

\item[\textbf{Taxa}]

@@

\item[\textbf{Characters}]
A list of characters used to distinguish between taxa. These are the questions that are to be presented to the user. A character can be categorical or numerical. When categorical, it has a list of states that are the relevant alternatives for this character.
\item[\textbf{Statements}]
Elements connecting a taxon to a character state. This is the core knowledge captured by the key: which taxa have which states or numerical values for which characters, thus distinguishing them from one another.

@@

The scientific name*, author, vernacular name, media elements, rank of a taxon. Info on whether it is a sub-taxonomic unit, and if it serves as an end-point or not. It can have a set of children, which are the underlying taxa. An external reference can define where info is to be retrieved from. It may have a geographic distribution as an object or external service, to assess where it occurs. A follow-up key as a url or external service reference can provide info on where a more detailed identification can be done.

@@

The name* and states* describing a property a taxon can have. It can have media elements and descriptions clarifying the character, and a user requirement. A logical premise can specify what other user input has to be given before the character may be presented. A character can be of the types ``exclusive'' (default, multiple choice where options exclude one another), ``non-exclusive'' (multiple choice with multiple answers possible), and ``numerical'' (the answer is a number). A character has states defining the possible answers.

@@

A possible answer for a character. Has either a title if the character type is non-numerical, or a min, max, step size and unit if it is numerical. It can have multiple media elements and descriptions clarifying the state. 

@@

A media element contains the data needed to display multimedia files. It can refer to different versions for different languages, and can have different versions for different media dimensions. It contains the required metadata such as width and height (images and video), length (sound and video), as well as creators, contributors, publishers and a license. Files can be urls to where the correct version of the file is to be found, or directly contain a base64 or svg encoded file. It supports external services to retrieve such information from elsewhere.

@@

The core element of any key is the statement. It defines a property of a taxon, separating it from other taxa with conflicting statements. In JSON (object) notation, a single statement takes the following form, in this case stating that Pikachu is an electric type Pokémon.

\begin{verbatim}
{
   "id": "statement:pikachu_is_electric",
   "taxonId": "taxon:pokemon_025",
   "characterId": "type_of_pokemon",
   "value": "state:pokemon_type_electric",
   "frequency": 1
}

@@

\subsection*{
Key metadata
}
Very few parameters are required on the top level of the key. Apart from the content of the key (taxa, characters and statements linking the two), a key is expected to refer to the version of the schema with which it complies, a title, language, license, a creator, the date at which it was last modified and an identifier that is to be kept stable across versions. As the creator is expected to be a reference to a person entity, one also needs to define at least one person.

@@

\subsection*{
Taxa
}
Eliminating all but one taxon is the goal of the user, and taxa are the units that all characteristics are connected to. Taxa can be provided as a simple (flat) list, but they can also be structured hierarchically, commonly, but not necessarily, adhering to their phylogeny. Statements can in such a hierarchy be connected to higher taxa, reducing a lot of the repetition of traits shared within a taxon that one would get in a simple list of taxa. If a statement is tied to a higher taxon in the hierarchy, it is implied that all the underlying taxa share the same statement.

In addition to biological taxonomic units, one can also define sub-groups within a taxon. Examples can be different sexes, morphs, or species complexes. Contrary to regular taxa, such subdivisions of taxa are not standalone taxonomic units. As such, they do not have their own scientific name but rather a label that adds specificity to their parent taxon. For example, the label of a sub-group specifying the sex of Pikachu would simply be ``\Female'' rather than ``Pikachu \Female'' as it already relates to the parent ``Pikachu''. For a default form of a taxon, the label can be an empty string.

The goal of the key is to provide a way to eliminate all but one taxon; the end point of the key.  The key should stop asking questions once the user has narrowed down the possible outcomes to a single end point. In a flat list of taxa, every taxon in the list will normally be an end point. In a hierarchical list, the lowermost taxa in the hierarchy will be the end points by default. This can be overridden, however, by explicitly tagging a taxon higher up the tree as an end point. Then, information on lower taxa may be displayed when identified while using the key, but no additional questions are asked once the end point has been determined. This can be used, for instance, to display an image of the relevant sub-group for a taxon, so that the taxon images reflect the input from the user.

\textit{\textbf{Example}: Pichu, Pikachu and Raichu are all species within the Pikachuidae. They each have a default and a ``Shiny'' morph, which is to be determined. The default Pikachu morph also has a subdivision into the sexes, but as the morph is tagged as an endpoint the user will not be asked further questions to determine these once the morph has been determined. See lines 25 - 94 in the \nameref{S2_Appendix} for the corresponding JSON, and Fig.~\ref{fig3} for a graphical representation.}

\begin{figure}[!h]
  \includegraphics[width=\textwidth]{Images/Fig3}
  \caption{\textit{\textbf{A taxon tree of the Pikachu-line}:
Taxa tagged as endpoint are indicated with an asterisk (*). This key continues until the correct morph (``Shiny'' or default) is known, while differences in the sexes are optional and will not be asked about once the correct Pokémon and morph have been determined.
  }}
  \label{fig3}
\end{figure}
\subsection*{
Characters, statements and frequencies
}
One of the possible values for a statement (the relationship between a taxon and a character) is the id of a state. This statement also may have a frequency: the proportion of cases where individuals of this taxon have this state for this character. It can be set to 0 or 1 or any value in between.

>>>>

\end{flushleft}
\newpage
\section*{
Abstract
}
The skills and knowledge needed to recognize and classify taxa is becoming increasingly scarce in the scientific community. At the same time, it is clear that these skills are strongly needed in biodiversity monitoring for management and conservation, especially when carried out by citizen scientists. Formalizing the required knowledge in the form of digital identification keys is one way of making such knowledge more available for professional and amateur observers of biodiversity. In this paper we describe Clavis, a modern open format for capturing knowledge required for taxon identification through digital keys, allowing for a level of detail beyond that of any current key format. We exemplify each concept using Pokémon as a fictional taxonomic group.
\section*{
Introduction
}
Distinguishing biological taxa from one another is a necessity in biodiversity monitoring. As species are going extinct at an unprecedented rate~\cite{Ceballos2015, Johnson2017}, we need to monitor as much of nature as we can to identify population statuses and trends. Meanwhile, research and management is currently facing a ``taxonomic impediment'', where taxonomic knowledge is gradually disappearing from the scientific community~\cite{Engel2021}.

Paradoxically, while taxonomic expertise is becoming scarce, species observations are being reported like never before. The bulk of the observational data currently available originates from citizen science, however, in which observations are made by non-professional volunteers~\cite{Silvertown2009}. There are large taxonomic biases in the data collected from these sources. For example, of publicly available data from the Global Biodiversity Information Facility (the largest aggregator of biodiversity observations in the world~\cite{GBIFhomepage}), some 67\% of over 2 billion observations pertain to birds~\cite{GBIF_data_taxonomy}. This means that less charismatic species, and species that are more difficult to identify, generally remain severely under-reported~\cite{Troudet2017}. Many such species play important ecological roles or can serve as indicators of broader ecosystem health, but the fact that few observers are able to reliably identify these species leads to their underreporting, thus obscuring important trends from the view of science and management~\cite{Dobson2020}.

The knowledge needed to identify species resides in large part in the heads of taxonomic experts. This invaluable experience is slowly disappearing as newly educated biologists are, to an increasing extent, trained with focus on skills in genetics, data management, and biodiversity informatics rather than traditional taxonomy. This training, and an academic reality of shorter periods with funding and more temporary contracts, gives fewer researchers the opportunity to invest the years of experience needed to become truly knowledgeable with regards to any substantial taxon, as was more commonplace in the past~\cite{Ebach2011}.

Identification keys are one of the most important tools used by taxonomists to identify taxa, as well as to share with others the information needed for distinguishing taxa. No taxonomist studying a large taxon in depth can get around identification keys, and keys are often the starting point when trying to learn a new organism group. Identification keys can, however, be quite challenging to use, especially for novel users, and their use often represents a significant barrier for the prospective learner. For this and other reasons, the use of identification keys is receiving less focus in most current biology curricula. Furthermore, as taxonomic knowledge is becoming increasingly fragmented, existing identification keys in literature are gradually becoming outdated as taxonomies are changing according to new insights.

These issues are exacerbated in the context of citizen science. Whereas prospective taxonomists have a professional incentive to learn to use a key, a steep learning curve can quickly put off the more casual citizen scientist. Citizen science, with its unrivaled amounts of data, but also added concerns regarding the reliability of the taxon identifications~\cite{Crall2011,Burgess2017} and the pronounced over-representation of more charismatic taxa~\cite{Callaghan2020,Bayraktarov2019,Boakes2016}, stands especially much to gain from good identification keys with a low user threshold.

Digital identification keys address these challenges in a number of ways. The formalized and machine-readable way in which the data are stored allows for easier revision, as well as multiple options for their display to the end user, tailored to their level of experience. Digital keys allow for the representation of more complex relationships between species' characteristics and identifications inherent to biological complexity. These relationships cannot as easily be represented in a linear form as is used in traditional, paper-based keys. With a suitable interface, digital keys open up the cumulative experience from taxonomic experts to a broader public like citizen scientists, while potentially being more reliable and educational than other identification methods such as automated image recognition. In contrast to paper-based keys, digital keys can also be combined with one another, integrating the information from several separate keys into one seamless user experience. And conversely, a digital key can easily be limited to just a subset of the taxa in it, for instance only taxa belonging to a certain habitat or geographic range. Another benefit is the possibility of saving the user input and the key together with the observational data, as a means of transparent identification that can be reviewed at any time in the future for quality control or in the case of new taxonomical insights.

Having a fully open and well-defined, platform-independent format is essential in ensuring that identification keys and the tools needed to display and create them remain both interoperable, interpretable across platforms, and freely available to all. A number of digital identification key formats exist~\cite{Dallwitz1980, Lucid}, but these come with a number of limitations in what they can represent, their ease of use, openness, etc. A well-defined format addressing this alleviates the technical burden in capturing taxonomic knowledge for future use.

In this manuscript we document Clavis; an open format for identification keys, aiming to cover the aforementioned requirements in an open and lightweight manner, serving as a way to store and exchange crucial taxonomic knowledge. Clavis is intended to capture all the requirements of traditional keys while adding the flexibility and complexity possible with digital keys. This means that the format is well suited for representing most, if not all, existing keys, both paper-based and digital, as well as for designing new keys. The name Clavis means ``key'' in Latin, and is a recursive acronym for \underline{C}lavis \underline{L}ightweight \underline{A}nd \underline{V}ersatile \underline{I}dentification \underline{S}chema. This article describes the Clavis format itself, example code handling the business logic of keys described using Clavis will be published separately at a later time. Descriptions of each of Clavis' data types and emerging properties are illustrated here using fictional taxa that are discretely defined and not subject to taxonomic debate or change, while exhibiting the required complexity for a demonstration of the more intricate features of the format. Our fictional creatures of choice for these examples are a selection of Pokémon~\cite{Pokemon_wiki}.

\section*{
Material and methods
}
Identification keys exist in many forms, but all are built on the same basic principle. The user is asked a number of questions about the entity being identified, and by answering these questions, taxa are excluded from consideration until (ideally) only one taxon remains, which is the result of the identification process.

Traditional single access keys consist of a decision tree with a single fixed path from beginning to end for each possible outcome. On choosing an answer from the alternatives for each of the questions, the user is directed to the next question to be answered, until they end up with a result taxon. Such keys can be either dichotomous, meaning that each question always has only two alternatives, leading to a bifurcating decision tree, or they can be multichotomous, meaning that each question may have more than two alternatives. One upside of single access keys is that they are easy to represent on paper. A drawback is that a user has to go through all questions in a specific order, having one and only one question to answer at any given time. This means that one cannot use only part of the key for any subset of taxa, nor easily go back and redo a question. Also, if a question cannot be answered, there is no way to proceed.

An alternative approach is the multiple access key. Such keys are typically stored as matrices where taxa and questions (characters) are stored as rows and columns in a tabular format, with cell values linking the taxa to their characters. Each character may have two or more possible answers (states). In a fully populated matrix key, every character is assigned a value (i.e. is scored) for every taxon. The user then has access to all the questions/characters at once and is not required to answer them in any particular order. Choosing any alternative for any of the questions will exclude from consideration all taxa that are not compatible with that alternative, bringing the user a step closer to the answer. The upside of this approach is that the user has several paths to the answer, and can choose to avoid questions that are difficult or impossible to answer. The downside is that the number of choices can easily be overwhelming, and it is in large part left to the user to try to choose the best path. Also, in real life situations it is very rare to find a set of characters that are both possible to score in a meaningful way for all the taxa in the key, and sufficient for distinguishing between them. More often, only some characters will be possible to score for all taxa, whereas others may be essential for distinguishing certain taxa in the key, yet be inapplicable to others.

To address this, a matrix can be filled out sparsely instead, meaning that not all characters are scored for all taxa. The interface can then display only those characters that are in fact scored for all the taxa currently under consideration. As taxa are being excluded by the user’s choices, further characters that are scored for all the remaining taxa will become available for answering. With this approach, it is possible to make a matrix key that behaves equivalently to a single access key. But the key can also be made to contain several distinguishing characters for any set of taxa, so that the user has more than one choice at each step and is thus not restricted to having to answer only one question at a time in a specific order.

The sparse matrix approach is more powerful than both the single access and the full matrix approach, in that both of these approaches are subsets of what the sparse matrix can represent. An important advantage of either matrix approach over the single access key is that the key can easily be restricted to only a subset of the taxa. This means that the user will not need to traverse the whole key if they wish to exclude one or more taxa a priori. Still, there is much additional information one would ideally like to store about taxa, characters, and the relationships between the two, which cannot be easily represented in a tabular format. For instance, the tabular format is not well suited for dealing with polymorphism in a straightforward way, i.e. the situation where a taxon can have more than one possible value for a given character. Also, it may often be desirable to treat taxa hierarchically rather than as a simple list of equally ranked entities, something which is difficult to do in the matrix format. Thus, representing taxonomic knowledge in this tabular format restricts the complexity of the information that can be stored.

The method of storing and exchanging taxonomic knowledge described here, is developed as a non-tabular multiple access key. The core of any Clavis key is a collection of statements. A statement links a taxon to a character, and specifies a state or numerical value the taxon has for that character (see Fig.~\ref{fig1}). E.g. the color (character) of species x (taxon) is red (state), or the length in millimeters (character) of genus x (taxon) is 4-11 (numerical range).

\begin{figure}[!h]
  \includegraphics[width=\textwidth]{Images/Fig1}
  \caption{{\bf The core concepts of any Clavis key.}
  \underline{Taxa} (left) and \underline{characters} (right) are connected through \underline{statements} (center). Each statement refers to one taxon and one character, linking the two by a value, usually one of the possible \underline{states} of the character, or a numerical value or range.}
  \label{fig1}
\end{figure}

Additional information can be linked to taxa, characters, states, and statements in a number of ways, capturing knowledge and relationships that cannot be represented in a tabular form in a practical way. Such information includes media elements, textual descriptions, and contextual information such as geographic scope. The Clavis format incorporates all the possibilities of the previously described approaches, so that any single access or matrix key can be transcribed to it.

Clavis compliant keys are written in JSON (JavaScript Object Notation)~\cite{JSON}, which has become the \textit{de facto} standard for data exchange over the internet, as it is an open, flexible format with widespread use and support within all modern programming languages. Clavis itself is a JSON-schema~\cite{JSON_schema}, a formal definition of the structure and content of a valid Clavis JSON file. JSON-schemas are machine readable and can be used to automatically validate the compliance of a JSON file in modern code editors, highlighting any issues that need to be solved. While it is possible to manually write the JSON of a Clavis-compliant key, it is generally not practical. In order to facilitate key creation and editing, one would generally provide taxonomists with a key editing interface that lets them easily record their knowledge, and let the key editor generate the corresponding JSON.
\subsection*{
Implementation
}
As an exchange and storage format, Clavis does not dictate how it is to be implemented. Different interfaces can implement it differently, depending on the purpose of the interface and the intended user group. To this end, interfaces serving to edit or display keys may disregard certain non-essential functionality supported by Clavis. One could for example create a key editor not supporting media files or multilingualism. Other aspects of the format are crucial however, and require support and unambiguous interpretation.

Identification is a matter of excluding taxa, and is done by letting the user select a state or numerical value for characters. Each time the user provides a new fact in this way, all taxa with conflicting statements are excluded (see Fig.~\ref{fig2}). The user can also be given the option to exclude a state rather than affirm one if there are more than two states to choose from.

@@

Whenever taxa are excluded, characters that were previously hidden may now have become relevant if all the remaining taxa have a statement pertaining to it, as illustrated in Fig.~\ref{fig2}. These characters should then be made visible to the user. The act of excluding taxa will also affect which of the possible states of a given character are relevant, even if it has not been answered directly by the user. If all the remaining taxa are known not to have a certain state, that state can be disabled. In such cases, the state still needs to be visible to the user as it provides context to the remaining alternatives, but it should not be possible for the user to select it. Characters that have not been answered, but that only have a single possible answer for all remaining taxa, can be hidden. In all cases, only characters that are linked by statements to \underline{all} the currently non-excluded taxa should be shown to the user. Clavis also allows for the inclusion of explicit dependencies, where the relevance of a character is determined by the answers given on one or more other characters.

Things become more complicated when allowing the user to undo previous answers. Undoing an answer might render a character that was subsequently answered irrelevant again. To avoid answers to this now irrelevant character from eliminating taxa as possibilities, an implementation needs to be able to remove or ignore answers on characters that have been rendered irrelevant.

@@

To illustrate the features supported by the Clavis format, we here describe features and give some example uses. The examples relate to fictional creatures that are not subject to taxonomic debate or change. For this we have chosen Pokémon as they appear in the mobile game Pokémon Go, as these are clearly defined, yet exhibit the complexity required to demonstrate the various features of the format. Keys for natural taxa would rarely use as much of the possible functionality as we demonstrate here in a single key. The key is valid in accordance with the JSON-schema, but it does not contain all the information needed to distinguish between all of the Pokémon mentioned in it. We also provide an example of a non-artificial key that does identify all taxa in it, but that does not aim to demonstrate all possible features that Clavis supports. This key covers all the species of titmice in Norway and can be found in the \nameref{S3_Appendix}.

@@

Polygon and/or name of the region where the key is valid

@@

The language(s) the key supports

@@

\item[\textbf{Taxa*}]

@@

\item[\textbf{Characters*}]
A list of characters used to distinguish between taxa. These are the questions that are presented to the user. A character can be categorical or numerical. When categorical, it has a list of states that are the relevant alternatives for this character.
\item[\textbf{Statements*}]
Elements connecting a taxon to a character through a state or numeric value. This is the core knowledge captured by the key: which taxa have which states or numerical values for which characters, thus distinguishing them from one another.

@@

The scientific name, author, vernacular name, label, media elements, rank of a taxon. Info on whether it serves as an end-point or not. It can have a set of children, which are the underlying taxa. An external reference can define where info is to be retrieved from. It may have a geographic distribution as an object or external service, to assess where it occurs. A follow-up key as a url or external service reference can provide info on where a more detailed identification can be done.

@@

The name* and states describing a property a taxon can have. It can have media elements and descriptions clarifying the character, and a user requirement. A logical premise can specify what other user input has to be given before the character may be presented. A character can be of the types ``exclusive'' (default, multiple choice where options exclude one another), ``non-exclusive'' (multiple choice with multiple answers possible), and ``numerical'' (the answer is a number). A character has states defining the possible answers if it is not numerical, and a min, max, unit and step size if it is numerical.

@@

A possible answer for a non-numerical character. Has a title and can have multiple media elements and descriptions clarifying the state. 

@@

A media element contains the data needed to display multimedia files. It can refer to different versions for different languages, and can have different versions for different media dimensions. It contains the required metadata such as width and height (images and video), length (sound and video), as well as creators, contributors, publishers and a license. Files can be urls to where the correct version of the file is to be found, or directly contain a base64 or svg encoded file. It supports external services to retrieve data from elsewhere.

@@

The core element of any key is the statement. It defines a property of a taxon, separating it from other taxa that have conflicting statements. In JSON code, a single statement takes the following form, in this case stating that Pikachu is an electric type Pokémon.

\begin{verbatim}
{
   "id": "statement:pikachu_is_electric",
   "taxon": "taxon:pokemon_025",
   "character": "character:type_of_pokemon",
   "value": "state:pokemon_type_electric",
   "frequency": 1
}

@@

\subsection*{
Key metadata
}
Very few parameters are required on the top level of the key. Apart from the content of the key (taxa, characters, and statements linking the two), a key is expected to refer to the version of the schema with which it complies, a title, language, license, a creator, the date at which it was last modified and an identifier that is to be kept stable across versions. As the creator has to be a reference to a person entity, at least one person needs to be defined as well.

@@

\subsection*{
Taxa
}
Eliminating all but one taxon is the goal of the user, and taxa are the units that all characteristics are connected to. Taxa can be provided as a flat list, but they can also be structured hierarchically, commonly, but not necessarily, adhering to their phylogeny. Statements can in such a hierarchy be connected to higher taxa, reducing a lot of the repetition of traits shared within a taxon that one would get when using a flat list of taxa. If a statement is tied to a higher taxon in the hierarchy, it is implied that all the underlying taxa share the same statement.

In addition to biological taxonomic units, one can also define sub-groups within a taxon. Examples can be different sexes, morphs, or species complexes. Contrary to regular taxa, such subdivisions of taxa are not standalone taxonomic units. As such, they do not have their own scientific name but rather a label that adds specificity to their parent taxon. For example, the label of a sub-group specifying the sex of Pikachu would simply be ``\Female'' rather than ``Pikachu \Female'' as it already relates to the parent taxon ``Pikachu''. For a default form of a taxon, the label can be an empty string.

The goal of the key is to provide a way to eliminate all but one taxon; the result of the identification process.  The key should stop asking questions once the user has narrowed down the possible outcomes to a single endpoint. In a flat list of taxa, every taxon in the list will be an endpoint. In a hierarchical list, the lowermost taxa in the hierarchy will be the endpoints by default. This can be overridden, however, by explicitly tagging a taxon higher up the tree as an endpoint. In this case, information on lower taxa may be displayed if identified while using the key, but no additional questions are asked once the endpoint has been determined. This can be used, for instance, to display an image of the relevant sub-group for a taxon, so that the taxon images reflect the input from the user.

\textit{\textbf{Example}: Pichu, Pikachu and Raichu are all species within the Pikachuidae. They each have a default and a ``Shiny'' morph, and in this key, the morph is the endpoint. The default Pikachu morph also has a subdivision into the sexes, but as the morph is tagged as an endpoint, the user will not be asked further questions to determine the sex once the morph has been determined. See lines 25 - 94 in the \nameref{S2_Appendix} for the corresponding JSON, and Fig.~\ref{fig3} for a graphical representation.}

\begin{figure}[!h]
  \includegraphics[width=\textwidth]{Images/Fig3}
  \caption{\textit{\textbf{A taxon tree of the Pikachu-line}:
Taxa tagged as endpoints are indicated with an asterisk (*). This key continues until the correct morph (``Shiny'' or default) is known. As the morph of Pikachu is defined as an endpoint, no questions regarding the sexes of Pikachu will be asked once the correct Pokémon and its morph have been determined.
  }}
  \label{fig3}
\end{figure}
\subsection*{
Characters, statements and frequencies
}
One of the possible values for a statement (the relationship between a taxon and a character) is the id of a state. A statement also may have a frequency: the proportion of cases where individuals of this taxon have this state for this character. It can be set to 0 or 1 or any value in between.
